Tropical mathematics and the Lambda-calculus I

### 2023-07-29 CSL24/section2bisbis.tex
commit

Let us start with a very basic probabilistic language:
the terms are $M::= \true \mid \false \mid M\oplus_p M$, for $p\in[0,1]$, and the operational semantics is $M\oplus_p N\to pM$ and $M\oplus_p N \to (1-p)N$, so that $M\oplus_p N$ plays the role of a probabilistic coin toss of bias $p$.
Consider the program
$$
 M:=(\true \oplus_p\false)\oplus_p((\true\oplus_p\false)\oplus_p(\false\oplus_p\true)).
 $$
 Calling $q=1-p$, to each occurrence of $\true$ or $\false$ in $M$, univocally determined by an address
$\omega\in \{l,r\}^{*}$, is associated a monomial $P_{\omega}(p,q)$ which determines the probability of the event ``$M\twoheadrightarrow_{\omega} \true/\false$'', that is, that $M$ reduces to $\true/\false$ according to the choices in $\omega$.
 For instance, we have
$P_{rll}(p,q):=qp^2$,
$P_{rrr}(p,q):=q^3$, and 
$P_{rrl}(p,q)=P_{rlr}(p,q):=q^2p$.
The polynomial function $Q_{\true}(p,q):=P_{ll}(p,q)+P_{rll}(p,q)+P_{rrr}(p,q)=p^2+p^2q+q^3$ gives instead the probability of the event ``$M\twoheadrightarrow \true$'', and analogously for $Q_{\false}(p,q):=P_{lr}(p,q)+P_{rrl}(p,q)+P_{rlr}(p,q)=pq+2pq^2$.

Two natural questions at this point concern the \emph{likelihood} of a given reduction address $\omega_{0}$:
%
%The tropical point of view allows now to express two natural questions about this situation:
\begin{enumerate}
 \item Which is the \emph{maximum likelihood estimator} for the event ``$M\twoheadrightarrow \false_{\omega_0}$''?
 I.e., which is the choice of $p,q$ that maximizes the probability $P_{\omega_0}$?
 \item 
Knowing that $M$ produced $\true$ (similarly for $\false$) , which is the \emph{maximum likelihood estimator} for the event ``$M\twoheadrightarrow \false_{\omega_0}$'', knowing that ``$M\twoheadrightarrow \false$''?
I.e., which is the choice of $p,q$ that makes $\omega_0$ the most likely path among those leading to $\false$ (i.e.\ that maximizes the conditional probability $\BB P(``M\twoheadrightarrow \false_{\omega_0}'' \mid ``M\twoheadrightarrow \false'')$)?
\end{enumerate}

>>>>

Let us start with a very basic probabilistic language:
the terms are $M::= \true \mid \false \mid M\oplus_p M$, for $p\in[0,1]$, and the operational semantics is $M\oplus_p N\to pM$ and $M\oplus_p N \to (1-p)N$, so that $M\oplus_p N$ plays the role of a probabilistic coin toss of bias $p$.
Consider the program
$$
 M:=(\true \oplus_p\false)\oplus_p((\true\oplus_p\false)\oplus_p(\false\oplus_p\true)).
 $$
 Calling $q=1-p$, to each occurrence of $\true$ or $\false$ in $M$, univocally determined by an address
$\omega\in \{l,r\}^{*}$, is associated a monomial $P_{\omega}(p,q)$ which determines the probability of the event ``$M\twoheadrightarrow_{\omega} \true/\false$'', that is, that $M$ reduces to $\true/\false$ according to the choices in $\omega$.
Thinking of $p,q$ as parameters, $P_{\omega}(p,q)$ can thus be read as the \emph{likelihood function} of the event ``$M\twoheadrightarrow \true_\omega/\false_\omega$''.
 For instance, we have
$P_{rll}(p,q):=qp^2$,
$P_{rrr}(p,q):=q^3$, and 
$P_{rrl}(p,q)=P_{rlr}(p,q):=q^2p$.
The polynomial function $Q_{\true}(p,q):=P_{ll}(p,q)+P_{rll}(p,q)+P_{rrr}(p,q)=p^2+p^2q+q^3$ gives instead the probability of the event ``$M\twoheadrightarrow \true$'', and analogously for $Q_{\false}(p,q):=P_{lr}(p,q)+P_{rrl}(p,q)+P_{rlr}(p,q)=pq+2pq^2$.
This way, the probabilistic evaluation of $M$ is presented as a \emph{hidden Markov model} \cite{Baumr966}, a fundamental statistical model, and notably one to which tropical methods are generally applied \cite{Pachter2ll4}.
Typical questions in this case would be, for a fixed $\omega_0$:
%
%The tropical point of view allows now to express two natural questions about this situation:
\begin{enumerate}
 \item Which is the \emph{maximum likelihood estimator} for the event ``$M\twoheadrightarrow \false_{\omega_0}$''?
 I.e., which is the choice of $p,q$ that maximizes the probability $P_{\omega_0}$?
 \item 
Knowing that $M$ produced $\true$ (similarly for $\false$) , which is the \emph{maximum likelihood estimator} for the event ``$M\twoheadrightarrow \false_{\omega_0}$'', knowing that ``$M\twoheadrightarrow \false$''?
I.e., which is the choice of $p,q$ that makes $\omega_0$ the most likely path among those leading to $\false$ (i.e.\ that maximizes the conditional probability $\BB P(``M\twoheadrightarrow \false_{\omega_0}'' \mid ``M\twoheadrightarrow \false'')$)?
\end{enumerate}
